---
title: "DS 6371 Project - House Prices Analysis"
author: "Jonathan Rocha & Samson Akomolafe"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cerulean
    highlight: tango
    toc: true
    toc_depth: 2
    toc_float: true
    toc_collapsed: true
    code_folding: show
runtime: shiny
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Load required libraries
library(shiny)
library(shinyjs)
library(shinythemes)
library(tidyverse)
library(DT)
library(ggplot2)
library(plotly)
library(ggthemes)
library(ggrepel)
library(car)       # For regression diagnostics
library(caret)     # For model training and evaluation
library(leaps)     # For model selection
library(MASS)      # For stepwise regression
library(lmtest)    # For model testing
library(glmnet)    # For regularized regression
```

# Introduction

This analysis explores the Ames Housing dataset, which contains
information on residential homes in Ames, Iowa. The dataset includes 79
explanatory variables describing various aspects of the homes, with the
goal of predicting the final sale price of each property.

Our analysis is divided into two main parts:

1.  **Analysis 1**: Examining how house sale prices relate to living
    area (GrLivArea) in three specific neighborhoods (NAmes, Edwards,
    and BrkSide) for Century 21 Ames.

2.  **Analysis 2**: Building predictive models for house prices across
    all neighborhoods in Ames, Iowa.

# Data Description

The Ames Housing dataset was compiled by Dean De Cock for use in data
science education. It contains information on residential homes in Ames,
Iowa, with 79 explanatory variables describing almost every aspect of
the properties.

-   **Source**: Kaggle competition "House Prices - Advanced Regression
    Techniques"
-   **Training set**: 1,460 observations
-   **Test set**: 1,459 observations
-   **Target variable**: SalePrice (the property's sale price in
    dollars)

```{r load_data}
# Load the datasets
train <- read.csv("~/Desktop/DS_6371_Project/house-prices-advanced-regression-techniques/train.csv")
test <- read.csv("~/Desktop/DS_6371_Project/house-prices-advanced-regression-techniques/test.csv")

# Display basic information about the training dataset
cat("Training set dimensions:", dim(train)[1], "rows,", dim(train)[2], "columns\n")
cat("Test set dimensions:", dim(test)[1], "rows,", dim(test)[2], "columns\n")
```

## Data Exploration

Let's first explore the structure of the data and get a summary of the
key variables.

```{r data_structure}
# Check the structure of the training dataset
str(train[, 1:10])  # Showing only the first 10 columns for brevity
```

```{r data_summary}
# Summary of key variables
summary(train[, c("SalePrice", "GrLivArea", "FullBath", "Neighborhood")])
```

```{r missing_values}
# Check for missing values in the training set
missing_values <- colSums(is.na(train))
missing_values[missing_values > 0]
```

## Data Visualization

Let's visualize the distribution of the target variable (SalePrice) and
its relationship with the living area (GrLivArea).

```{r price_distribution}
# Distribution of SalePrice
ggplot(train, aes(x = SalePrice)) +
  geom_histogram(fill = "steelblue", bins = 30) +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "Distribution of Sale Prices",
       x = "Sale Price ($)",
       y = "Frequency") +
  theme_minimal()
```

```{r price_vs_area}
# Relationship between SalePrice and GrLivArea
ggplot(train, aes(x = GrLivArea, y = SalePrice)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Sale Price vs. Living Area",
       x = "Above Grade (Ground) Living Area (sq ft)",
       y = "Sale Price ($)") +
  theme_minimal()
```

# Analysis 1: Century 21 Ames Neighborhood Analysis

## Restatement of Problem

Century 21 Ames, a real estate company in Ames, Iowa, has commissioned
us to analyze how the sale price of houses is related to the square
footage of the living area (GrLivArea) and whether this relationship
varies by neighborhood. The company only sells houses in three
neighborhoods: NAmes, Edwards, and BrkSide.

Our task is to: 1. Build a model that relates sale price to living area
2. Determine if this relationship differs by neighborhood 3. Provide
estimates and confidence intervals for the relationship 4. Ensure model
assumptions are met and address any outliers 5. Present the results in
terms of 100 sq. ft. increments of living area

```{r century21_data}
# Filter data for the three neighborhoods of interest
century21_data <- train %>%
  filter(Neighborhood %in% c("NAmes", "Edwards", "BrkSide"))

# Count of houses in each neighborhood
table(century21_data$Neighborhood)

# Basic statistics by neighborhood
century21_data %>%
  group_by(Neighborhood) %>%
  summarize(
    Count = n(),
    Mean_Price = mean(SalePrice),
    Median_Price = median(SalePrice),
    Mean_Area = mean(GrLivArea),
    Median_Area = median(GrLivArea)
  ) %>%
  knitr::kable(digits = 2)
```

```{r century21_visualization}
# Visualize the relationship between SalePrice and GrLivArea by neighborhood
ggplot(century21_data, aes(x = GrLivArea, y = SalePrice, color = Neighborhood)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Sale Price vs. Living Area by Neighborhood",
       x = "Above Grade (Ground) Living Area (sq ft)",
       y = "Sale Price ($)") +
  theme_minimal()
```

## Build and Fit the Model

We'll build several models to examine the relationship between sale
price and living area, considering the neighborhood effect:

1.  A simple linear regression model with GrLivArea as the predictor
2.  A model with GrLivArea and Neighborhood as predictors
3.  A model with interaction between GrLivArea and Neighborhood

```{r century21_models}
# Scale GrLivArea to 100 sq ft units for easier interpretation
century21_data$GrLivArea_100 <- century21_data$GrLivArea / 100

# Model 1: Simple linear regression
model1 <- lm(SalePrice ~ GrLivArea_100, data = century21_data)

# Model 2: Including neighborhood
model2 <- lm(SalePrice ~ GrLivArea_100 + Neighborhood, data = century21_data)

# Model 3: Including interaction
model3 <- lm(SalePrice ~ GrLivArea_100 * Neighborhood, data = century21_data)

# Summary of models
summary(model1)
summary(model2)
summary(model3)
```

## Checking Assumptions

We'll check the assumptions for the best model: 1. Linearity 2.
Independence 3. Normality 4. Equal variance (homoscedasticity)

```{r model_assumptions}
# Assuming model3 is the best model based on the results
par(mfrow = c(2, 2))
plot(model3)
```

## Influential Point Analysis

We'll identify and analyze potential outliers and influential points.

```{r influential_points}
# Calculate Cook's distance
cooksd <- cooks.distance(model3)

# Plot Cook's distance
plot(cooksd, pch = 19, main = "Cook's Distance", ylab = "Cook's Distance")
abline(h = 4/nrow(century21_data), col = "red")  # Threshold line

# Identify influential points
influential_points <- which(cooksd > 4/nrow(century21_data))
century21_data[influential_points, c("Id", "Neighborhood", "GrLivArea", "SalePrice")]
```

## Comparing Competing Models

We'll compare the models using various metrics:

```{r model_comparison}
# Function to calculate CV PRESS
cv_press <- function(model) {
  pr <- residuals(model)/(1 - hatvalues(model))
  sum(pr^2)
}

# Compare models
model_comparison <- data.frame(
  Model = c("Model 1: GrLivArea", 
            "Model 2: GrLivArea + Neighborhood", 
            "Model 3: GrLivArea * Neighborhood"),
  Adj_R2 = c(summary(model1)$adj.r.squared,
             summary(model2)$adj.r.squared,
             summary(model3)$adj.r.squared),
  CV_PRESS = c(cv_press(model1),
               cv_press(model2),
               cv_press(model3)),
  AIC = c(AIC(model1),
          AIC(model2),
          AIC(model3))
)

knitr::kable(model_comparison, digits = 4)
```

## Parameter Estimates and Interpretation

We'll provide estimates and confidence intervals for the parameters of
the best model.

```{r parameter_estimates}
# Confidence intervals for the best model
conf_int <- confint(model3, level = 0.95)
conf_int
```

## Conclusion for Analysis 1

[This section will be completed after running the analysis and
interpreting the results]

# Analysis 2: Predictive Modeling for All Neighborhoods

## Restatement of Problem

For this analysis, we need to build the most predictive model for sale
prices of homes in all of Ames, Iowa. We'll create at least three
competing models:

1.  A simple linear regression model
2.  A multiple linear regression model with GrLivArea and FullBath as
    predictors
3.  At least one additional multiple linear regression model with
    selected variables

We'll compare these models using adjusted R², CV PRESS, AIC, and Kaggle
Score.

```{r all_data_exploration}
# Explore relationships between SalePrice and key variables
par(mfrow = c(2, 2))
plot(train$GrLivArea, train$SalePrice, main = "SalePrice vs GrLivArea", 
     xlab = "GrLivArea", ylab = "SalePrice")
plot(train$FullBath, train$SalePrice, main = "SalePrice vs FullBath", 
     xlab = "FullBath", ylab = "SalePrice")
plot(train$YearBuilt, train$SalePrice, main = "SalePrice vs YearBuilt", 
     xlab = "YearBuilt", ylab = "SalePrice")
plot(train$OverallQual, train$SalePrice, main = "SalePrice vs OverallQual", 
     xlab = "OverallQual", ylab = "SalePrice")
```

## Candidate Models

### Simple Linear Regression

```{r slr_model}
# Simple linear regression with GrLivArea
slr_model <- lm(SalePrice ~ GrLivArea, data = train)
summary(slr_model)
```

### Multiple Linear Regression 1

```{r mlr1_model}
# Multiple linear regression with GrLivArea and FullBath
mlr1_model <- lm(SalePrice ~ GrLivArea + FullBath, data = train)
summary(mlr1_model)
```

### Multiple Linear Regression 2

For our third model, we'll select variables based on correlation with
SalePrice and domain knowledge.

```{r variable_selection}
# Calculate correlations with SalePrice
numeric_vars <- sapply(train, is.numeric)
correlations <- cor(train[, numeric_vars], train$SalePrice, use = "complete.obs")
correlations <- sort(abs(correlations), decreasing = TRUE)
head(correlations, 10)
```

```{r mlr2_model}
# Multiple linear regression with selected variables
mlr2_model <- lm(SalePrice ~ GrLivArea + OverallQual + YearBuilt + 
                  TotalBsmtSF + GarageCars, data = train)
summary(mlr2_model)
```

## Checking Assumptions

We'll check the assumptions for the best model.

```{r mlr2_assumptions}
# Check assumptions for MLR2 model
par(mfrow = c(2, 2))
plot(mlr2_model)
```

## Influential Point Analysis

```{r mlr2_influential}
# Calculate Cook's distance for MLR2
cooksd_mlr2 <- cooks.distance(mlr2_model)

# Plot Cook's distance
plot(cooksd_mlr2, pch = 19, main = "Cook's Distance", ylab = "Cook's Distance")
abline(h = 4/nrow(train), col = "red")  # Threshold line

# Identify influential points
influential_points_mlr2 <- which(cooksd_mlr2 > 4/nrow(train))
head(train[influential_points_mlr2, c("Id", "GrLivArea", "SalePrice")], 10)
```

## Comparing Competing Models

```{r all_model_comparison}
# Compare all models
all_model_comparison <- data.frame(
  Model = c("Simple Linear Regression", 
            "Multiple Linear Regression 1", 
            "Multiple Linear Regression 2"),
  Adj_R2 = c(summary(slr_model)$adj.r.squared,
             summary(mlr1_model)$adj.r.squared,
             summary(mlr2_model)$adj.r.squared),
  CV_PRESS = c(cv_press(slr_model),
               cv_press(mlr1_model),
               cv_press(mlr2_model)),
  AIC = c(AIC(slr_model),
          AIC(mlr1_model),
          AIC(mlr2_model)),
  Kaggle_Score = c(NA, NA, NA)  # To be filled after submission
)

knitr::kable(all_model_comparison, digits = 4)
```

## Kaggle Submission

We'll prepare predictions for the test set using our best model and
create a submission file for Kaggle.

```{r kaggle_submission, eval=FALSE}
# Predict on test set using the best model (MLR2)
predictions <- predict(mlr2_model, newdata = test)

# Create submission file
submission <- data.frame(
  Id = test$Id,
  SalePrice = predictions
)

# Write to CSV
write.csv(submission, "submission.csv", row.names = FALSE)
```

## Conclusion for Analysis 2

[This section will be completed after running the analysis and
interpreting the results]

# R Shiny App: Price vs. Living Area Chart

```{r shiny_app, echo=TRUE}
# Define UI
ui <- fluidPage(
  titlePanel("House Price vs. Living Area by Neighborhood"),
  
  sidebarLayout(
    sidebarPanel(
      checkboxGroupInput("neighborhoods", 
                         "Select Neighborhoods:",
                         choices = c("NAmes", "Edwards", "BrkSide"),
                         selected = c("NAmes", "Edwards", "BrkSide")),
      
      sliderInput("area_range", 
                  "Living Area Range (sq ft):",
                  min = min(century21_data$GrLivArea),
                  max = max(century21_data$GrLivArea),
                  value = c(min(century21_data$GrLivArea), 
                            max(century21_data$GrLivArea))),
      
      checkboxInput("show_regression", 
                    "Show Regression Line", 
                    value = TRUE),
      
      checkboxInput("show_confidence", 
                    "Show Confidence Interval", 
                    value = TRUE)
    ),
    
    mainPanel(
      plotlyOutput("scatterplot"),
      br(),
      tableOutput("neighborhood_stats")
    )
  )
)

# Define server logic
server <- function(input, output) {
  
  # Filter data based on user input
  filtered_data <- reactive({
    century21_data %>%
      filter(Neighborhood %in% input$neighborhoods,
             GrLivArea >= input$area_range[1],
             GrLivArea <= input$area_range[2])
  })
  
  # Create scatterplot
  output$scatterplot <- renderPlotly({
    p <- ggplot(filtered_data(), aes(x = GrLivArea, y = SalePrice, color = Neighborhood)) +
      geom_point(alpha = 0.7) +
      scale_y_continuous(labels = scales::comma) +
      labs(title = "Sale Price vs. Living Area",
           x = "Above Grade (Ground) Living Area (sq ft)",
           y = "Sale Price ($)") +
      theme_minimal()
    
    if (input$show_regression) {
      p <- p + geom_smooth(method = "lm", se = input$show_confidence)
    }
    
    ggplotly(p)
  })
  
  # Create neighborhood statistics table
  output$neighborhood_stats <- renderTable({
    filtered_data() %>%
      group_by(Neighborhood) %>%
      summarize(
        Count = n(),
        `Mean Price` = mean(SalePrice),
        `Median Price` = median(SalePrice),
        `Mean Area` = mean(GrLivArea),
        `Median Area` = median(GrLivArea),
        `Price per sq ft` = mean(SalePrice) / mean(GrLivArea)
      ) %>%
      arrange(desc(`Mean Price`))
  }, digits = 0)
}

# Run the application
shinyApp(ui = ui, server = server)
```

# Appendix: R Code

All the R code used in this analysis is included in the R Markdown
document.
